### 2023-03-02 chapters/conclusion.tex
minor fix: section -> chapter

\begin{itemize}
	\item 
	\textbf{Post-hoc Explainability.} Black-box classifiers are hard to understand and trust, especially in safety-critical domains. In this thesis, we have designed interpretable classifiers that are explainable by design. We can use these classifiers, such as {\imli} and {\crr}, to generate post-hoc explanations for black-box classifiers. In this context, logic-based explanations, such as abductive explanations, Anchor, LORE, CORELS etc., are effective for this purpose. Hence, we plan to test how well {\imli} can synthesize explanations with large samples and how well {\crr} can handle more expressive explanations.

	\item \textbf{Fairness Auditing.} Machine learning has grown as a predictive system in automated decision making. However, quantifying and mitigating bias in these systems is an active area of research. There has been significant research interest in auditing classifiers for bias, from standard supervised learning to deep neural networks, computer vision, large language models and so on. Thus, we need a fairness auditing framework with formal guarantees. One task of this framework is to categorize different definitions of fairness metrics in the literature. This would help stakeholders pick the right definition of fairness for their application. We plan to extend formal fairness verification beyond group fairness, such as individual fairness, causal fairness, and counterfactual fairness. We also plan to cover more classes of machine learning classifiers that are hard or impossible to encode as CNF, such as deep neural networks. Additionally, we aim to extend fairness verification beyond tabular data to more complex data such as images and texts.

	\item \textbf{Fairness and Interpretability: Bridging the Gap.} We want to make machine learning models fair and interpretable. We are interested in how these two goals relate to each other. We have some research questions for the future: (i) are interpretable models fair? (ii) how can we improve the fairness of an interpretable model? (iii) how can we use interpretability to enhance fairness? (iv) how can we optimize a classifier for both fairness and interpretability? Our work on {\fairXplainer} is a stepping stone towards bridging the gap between fairness and interpretability. We plan to explore more directions in the future.

	\item \textbf{Efficiency through Incremental Solving.} We will also apply incremental solving, which we used for interpretable classification learning with {\imli} and {\crr}, to other optimization problems, not only in machine learning. This can help us scale up without losing much accuracy. For example, we want to use incremental solving in online learning of interpretable models, such as decision trees, decision lists, and decision sets. To conclude, the long-term vision of this thesis is to enhance machine learning in interpretability, fairness, and beyond, with the help of formal methods and automated reasoning.
\end{itemize}

>>>>

\begin{itemize}
	\item 
	\textbf{Post-hoc Explainability.} Black-box classifiers are hard to understand and trust, especially in safety-critical domains. In this thesis, we have designed interpretable classifiers that are explainable by design. We can use these classifiers, such as {\imli} and {\crr}, to generate post-hoc explanations for black-box classifiers. In this context, logic-based explanations, such as abduction-based explanations~\cite{ignatiev2019abduction}, Anchor~\cite{ribeiro2018anchors}, and LORE~\cite{guidotti2018local} have demonstrated to be effective for this purpose. Therefore, we plan to test how well {\imli} can synthesize explanations with large samples and how well {\crr} can generate more expressive explanations.

	\item \textbf{Fairness Auditing.} Machine learning has demonstrated significant success as a predictive system in automated decision-making. However, quantifying and mitigating bias in these systems is still challenging, e.g. bias in generative machine learning and large language models~\cite{abid2021persistent,nadeem2020stereoset,vig2020investigating}. As a result, there has been significant research interest in auditing classifiers for bias, from standard supervised learning to deep neural networks, computer vision, large language models and so on. Thus, we need a fairness auditing framework~\cite{ruf2021towards,yan2022active} with formal guarantees. One task of this framework is to categorize different definitions of fairness metrics in the literature. This would help stakeholders pick the right definition of fairness for their application~\cite{ruf2021towards}. We plan to extend formal fairness verification beyond group fairness, such as individual fairness~\cite{john2020verifying}, causal fairness~\cite{pan2021explaining,zhang2018fairness}, and counterfactual fairness~\cite{wu2019counterfactual,chiappa2019path}. We also plan to cover more classes of machine learning classifiers that are hard or impossible to encode as CNF, such as deep neural networks. Additionally, we aim to extend fairness verification beyond tabular data to more complex data such as images and texts.

	\item \textbf{Fairness and Interpretability: Bridging the Gap.} We aim to design fair and interpretable algorithms for  machine learning. We are interested in how these two goals relate to each other. We have some research questions for the future: (i) are interpretable models fair? (ii) how can we improve the fairness of interpretable models? (iii) how can we use interpretability to enhance fairness? (iv) how can we optimize a classifier for both fairness and interpretability? Our work on {\fairXplainer} is a stepping stone towards bridging the gap between fairness and interpretability. We plan to explore more directions in the future.

	\item \textbf{Efficiency through Incremental Solving.} We aim to apply incremental solving, which we used for interpretable classification learning with {\imli} and {\crr}, to other optimization problems, not only in machine learning. This can help us scale up without losing much accuracy. For example, we aim to apply incremental solving in online learning of interpretable models, such as decision trees, decision lists, and decision sets. To conclude, the long-term vision of this thesis is to enhance machine learning in interpretability, fairness, and beyond, with the help of formal methods and automated reasoning.
\end{itemize}

### 2023-03-02 chapters/fairness/fif/appendix.tex
added appendix
minor fix: section -> chapter

\appendix
\chapter{Fairness Influence Functions}
%\section{Societal Impact}
%In this paper, we quantify the influence of input features on the incurred bias/unfairness of the classifier on a dataset. This quantification facilitates our understanding of potential features or subsets of features attributing highly to the bias. This also allows us to understand the effect of the fairness enhancing algorithms in removing bias and their achievement/failure to do so. To the best of our understanding, this paper does not have any negative societal impact.

@@

Next, we apply variance decomposition (Equation~\eqref{fairness_fairXplainer_eq:variance_decomposition_set_notation}) to estimate the influence of each $ \nonsensitive_{\mathbf{S}} $. From Equation~\eqref{fairness_fairXplainer_eq:variance_decomposition_set_notation} in Section~\ref{fairness_fairXplainer_sec:preliminaries}, we obtain that the variance of $ B_{\mathbf{a}} $ can be decomposed as 
\[ \mathsf{Var}[B_\mathbf{a}] = \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} V_{\mathbf{a}, \mathbf{S}},\] 
where $ V_{\mathbf{a}, \mathbf{S}} $ denotes the decomposed variance of $ B_\mathbf{a} $ w.r.t.\ the features $ \nonsensitive_{\mathbf{S}} $ conditioned on the sensitive group $ \sensitive = \mathbf{a} $.

@@

In this paper, we consider cubic splines with the basis function $ B_{r,4}(Z) $ that constitutes a B-spline of degree $ 3 $. This polynomial has $ C^2 $ continuity, i.e. for each piecewise term, derivatives up to the second order are zero at the endpoints of each interval in the knot vector. We estimate component functions $ \fiffunc_{\mathbf{a}, \mathbf{S}} $'s with the basis function $ B_{r,4}(\mathbf{Z})  $ of cubic splines~\cite{li2010global}, as shown in Equation~\eqref{fairness_fairXplainer_eq:cubic_splines}.  
\begin{align}\label{fairness_fairXplainer_eq:cubic_splines}
\begin{split}
&\fiffunc_{\mathbf{a}, \{i\}} (\mathbf{Z}_{\{i\}}) \approx \sum_{r=-1}^{\tau+1}\alpha_r^iB_{r,n}(\mathbf{Z}_{\{i\}})\\
&\fiffunc_{\mathbf{a}, \{i, j\}} (\mathbf{Z}_{\{i, j\}}) \approx \sum_{p=-1}^{\tau+1}\sum_{q=-1}^{\tau+1}\beta_{pq}^{ij}B_p(\mathbf{Z}_{\{i\}})B_q(\mathbf{Z}_{\{j\}})\\
&\fiffunc_{\mathbf{a}, \{i, j, k\}} (\mathbf{Z}_{\{i, j, k\}}) \approx \sum_{p=-1}^{\tau+1}\sum_{q=-1}^{\tau+1}\sum_{r=-1}^{\tau+1}\gamma_{pqr}^{ijk}B_p(\mathbf{Z}_{\{i\}})B_q(\mathbf{Z}_{\{j\}})B_{r,n}(\mathbf{Z}_{\{j\}})
\end{split}
\end{align}

@@

Following the classification setting and notations described in Section~\ref{chapter:preliminaries}, here we consider a binary classifier trained on dataset $\mathbf{D}\triangleq \{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$ as $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $. $\widehat{Y} \in \{0,1\}$ and ${Y} \in \{0,1\}$ represents the predicted class and the true class for a data point $ (\nonsensitive, \sensitive) $ with sensitive features $\sensitive$ and non-sensitive features $\nonsensitive$.

@@

The majority of dots for {\fairXplainer} are on the line connecting $ (0, 0) $ to $ (1, 1) $  in Figure~\ref{fairness_fairXplainer_fig:accuracy_scatter_plot}. This implies that the estimation of {\fairXplainer} is significantly accurate. Interestingly, error rarely occurs in {\fairXplainer} due to the degenerate cases mentioned in Section~\ref{fairness_fairXplainer_sec:fifs}. 
In contrast, SHAP often computes a metric inaccurately, thereby showing the drawback of applying a local explanation approach for FIF computation. \emph{Thus, {\fairXplainer} is significantly more accurate than SHAP in computing FIFs and the corresponding fairness metrics.}

@@

\subsection{FIF of Different Datasets}
We deploy a neural network ($ 3 $ hidden layers, each with $ 2 $ neurons, L$ 2 $ penalty regularization term as $ 10^{-5} $, a constant learning rate as $ 0.001 $) on different datasets, namely Adult and Titanic, and demonstrate the corresponding FIFs in Figures~\ref{fairness_fairXplainer_fig:individual_vs_intersectional_influence_adult} and \ref{fairness_fairXplainer_fig:individual_vs_intersectional_influence_titanic}, respectively. In all the figures, both individual and intersectional FIFs depict the sources of bias more clearly than individual FIFs alone, as argued in Section~\ref{fairness_fairXplainer_sec:experiments}.

>>>>

\chapter{Fairness Influence Functions}
%\section{Societal Impact}
%In this paper, we quantify the influence of input features on the incurred bias/unfairness of the classifier on a dataset. This quantification facilitates our understanding of potential features or subsets of features attributing highly to the bias. This also allows us to understand the effect of the fairness enhancing algorithms in removing bias and their achievement/failure to do so. To the best of our understanding, this paper does not have any negative societal impact.

@@

Next, we apply variance decomposition (Equation~\eqref{fairness_fairXplainer_eq:variance_decomposition_set_notation}) to estimate the influence of each $ \nonsensitive_{\mathbf{S}} $. From Equation~\eqref{fairness_fairXplainer_eq:variance_decomposition_set_notation} in Chapter~\ref{fairness_fairXplainer_sec:preliminaries}, we obtain that the variance of $ B_{\mathbf{a}} $ can be decomposed as 
\[ \mathsf{Var}[B_\mathbf{a}] = \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} V_{\mathbf{a}, \mathbf{S}},\] 
where $ V_{\mathbf{a}, \mathbf{S}} $ denotes the decomposed variance of $ B_\mathbf{a} $ w.r.t.\ the features $ \nonsensitive_{\mathbf{S}} $ conditioned on the sensitive group $ \sensitive = \mathbf{a} $.

@@

In this chapter, we consider cubic splines with the basis function $ B_{r,4}(Z) $ that constitutes a B-spline of degree $ 3 $. This polynomial has $ C^2 $ continuity, i.e. for each piecewise term, derivatives up to the second order are zero at the endpoints of each interval in the knot vector. We estimate component functions $ \fiffunc_{\mathbf{a}, \mathbf{S}} $'s with the basis function $ B_{r,4}(\mathbf{Z})  $ of cubic splines~\cite{li2010global}, as shown in Equation~\eqref{fairness_fairXplainer_eq:cubic_splines}.  
\begin{align}\label{fairness_fairXplainer_eq:cubic_splines}
\begin{split}
&\fiffunc_{\mathbf{a}, \{i\}} (\mathbf{Z}_{\{i\}}) \approx \sum_{r=-1}^{\tau+1}\alpha_r^iB_{r,n}(\mathbf{Z}_{\{i\}})\\
&\fiffunc_{\mathbf{a}, \{i, j\}} (\mathbf{Z}_{\{i, j\}}) \approx \sum_{p=-1}^{\tau+1}\sum_{q=-1}^{\tau+1}\beta_{pq}^{ij}B_p(\mathbf{Z}_{\{i\}})B_q(\mathbf{Z}_{\{j\}})\\
&\fiffunc_{\mathbf{a}, \{i, j, k\}} (\mathbf{Z}_{\{i, j, k\}}) \approx \sum_{p=-1}^{\tau+1}\sum_{q=-1}^{\tau+1}\sum_{r=-1}^{\tau+1}\gamma_{pqr}^{ijk}B_p(\mathbf{Z}_{\{i\}})B_q(\mathbf{Z}_{\{j\}})B_{r,n}(\mathbf{Z}_{\{j\}})
\end{split}
\end{align}

@@

Following the classification setting and notations described in Chapter~\ref{chapter:preliminaries}, here we consider a binary classifier trained on dataset $\mathbf{D}\triangleq \{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$ as $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $. $\widehat{Y} \in \{0,1\}$ and ${Y} \in \{0,1\}$ represents the predicted class and the true class for a data point $ (\nonsensitive, \sensitive) $ with sensitive features $\sensitive$ and non-sensitive features $\nonsensitive$.

@@

The majority of dots for {\fairXplainer} are on the line connecting $ (0, 0) $ to $ (1, 1) $  in Figure~\ref{fairness_fairXplainer_fig:accuracy_scatter_plot}. This implies that the estimation of {\fairXplainer} is significantly accurate. Interestingly, error rarely occurs in {\fairXplainer} due to the degenerate cases mentioned in Chapter~\ref{fairness_fairXplainer_sec:fifs}. 
In contrast, SHAP often computes a metric inaccurately, thereby showing the drawback of applying a local explanation approach for FIF computation. \emph{Thus, {\fairXplainer} is significantly more accurate than SHAP in computing FIFs and the corresponding fairness metrics.}

@@

\subsection{FIF of Different Datasets}
We deploy a neural network ($ 3 $ hidden layers, each with $ 2 $ neurons, L$ 2 $ penalty regularization term as $ 10^{-5} $, a constant learning rate as $ 0.001 $) on different datasets, namely Adult and Titanic, and demonstrate the corresponding FIFs in Figures~\ref{fairness_fairXplainer_fig:individual_vs_intersectional_influence_adult} and \ref{fairness_fairXplainer_fig:individual_vs_intersectional_influence_titanic}, respectively. In all the figures, both individual and intersectional FIFs depict the sources of bias more clearly than individual FIFs alone, as argued in Chapter~\ref{fairness_fairXplainer_sec:experiments}.

### 2023-03-02 chapters/fairness/fif/framework.tex
minor fix: section -> chapter

%\clearpage
\section{Fairness Influence Functions: Formulation and Properties}\label{fairness_fairXplainer_sec:fifs}
We formalize Fairness Influence Functions (FIF) as a quantifier of the contribution of a subset of features to the resultant bias of a classifier applied on a dataset.  In GSA, we observe that the variance of the output of a function can be attributed to the corresponding subset of input variables through variance decomposition (Eq.~\eqref{fairness_fairXplainer_eq:variance_decomposition_set_notation}). To leverage the power of GSA in fairness in machine learning, we first express the existing fairness metrics in terms of the variance of classifier's prediction in Section~\ref{fairness_fairXplainer_sec:metric_as_variance}. This allows us to formulate FIF by leveraging variance decomposition (Section~\ref{fairness_fairXplainer_sec:fif_formulation}).

@@

Informally, the FIF $ w_{\mathbf{S}} $ of $ \feature_{\mathbf{S}} $ is the difference in the scaled decomposed variance of CPPs between sensitive groups $ \mathbf{a}_{\max} $ and $ \mathbf{a}_{\min} $ as induced by  $ \feature_{\mathbf{S}} $. Thus, FIF of features is non-zero when the scaled decomposed variance-difference of CPPs is non-zero for those features, and vice versa. We refer to $ w_{\mathbf{S}} $ as an \emph{individual influence} when $ |\mathbf{S}| = 1 $, and as an \emph{intersectional influence} when $ |\mathbf{S}| > 1 $. Being able to naturally quantify the higher-order influences allow FIFs to interpret the sources of bias in a more fine-grained manner. We experimentally validate this in Section~\ref{fairness_fairXplainer_sec:experiments}.

@@

\textit{Expressing FIF in terms of the variance decomposition over a subset of features allows us to import and extend well-studied techniques of GSA to perform FIF estimation}, which we elaborate on Section~\ref{fairness_fairXplainer_sec:fairxplainer}.

>>>>

%\clearpage
\section{Fairness Influence Functions: Formulation and Properties}\label{fairness_fairXplainer_sec:fifs}
We formalize Fairness Influence Functions (FIF) as a quantifier of the contribution of a subset of features to the resultant bias of a classifier applied on a dataset.  In GSA, we observe that the variance of the output of a function can be attributed to the corresponding subset of input variables through variance decomposition (Eq.~\eqref{fairness_fairXplainer_eq:variance_decomposition_set_notation}). To leverage the power of GSA in fairness in machine learning, we first express the existing fairness metrics in terms of the variance of classifier's prediction in Chapter~\ref{fairness_fairXplainer_sec:metric_as_variance}. This allows us to formulate FIF by leveraging variance decomposition (Chapter~\ref{fairness_fairXplainer_sec:fif_formulation}).

@@

Informally, the FIF $ w_{\mathbf{S}} $ of $ \feature_{\mathbf{S}} $ is the difference in the scaled decomposed variance of CPPs between sensitive groups $ \mathbf{a}_{\max} $ and $ \mathbf{a}_{\min} $ as induced by  $ \feature_{\mathbf{S}} $. Thus, FIF of features is non-zero when the scaled decomposed variance-difference of CPPs is non-zero for those features, and vice versa. We refer to $ w_{\mathbf{S}} $ as an \emph{individual influence} when $ |\mathbf{S}| = 1 $, and as an \emph{intersectional influence} when $ |\mathbf{S}| > 1 $. Being able to naturally quantify the higher-order influences allow FIFs to interpret the sources of bias in a more fine-grained manner. We experimentally validate this in Chapter~\ref{fairness_fairXplainer_sec:experiments}.

@@

\textit{Expressing FIF in terms of the variance decomposition over a subset of features allows us to import and extend well-studied techniques of GSA to perform FIF estimation}, which we elaborate on Chapter~\ref{fairness_fairXplainer_sec:fairxplainer}.
